|

### 2025-05-29 00-abstract.Rmd
revised version

How does integrating large language models (LLMs) into classroom activities influence students' learning? To investigate this question, we conducted a randomized experiment with students aged 14-16 (N=122), testing two interventions: (i) comparing AI-generated solutions with step-by-step explanations versus solutions alone, and (ii) Socratic AI chatbots promoting critical thinking through incremental guidance versus non-Socratic AI offering direct solutions. Students completed school-related tasks under these conditions. We collected students' performance and interaction logs data, and their attitudes through self-reported surveys. Results indicated that AI-generated explanations significantly improved students' performance over solutions alone, highlighting the benefits of step-by-step guidance. The Socratic AI approach fostered significantly greater engagement and interaction. However, it did not achieve significant improvements in learning, and a higher fraction of students perceived it as less helpful. Furthermore, despite overall positive perceptions of AI assistance, students exhibited limited retention failing to apply learned concepts to new situations when AI assistance was removed. These findings underscore both the potential and limitations of LLM-based chatbots in k-12 education, especially in balancing critical thinking development with user satisfaction.

>>>>

How does integrating large language models (LLMs) into classroom activities influence students' learning? To investigate this question, we conducted a randomized experiment with students aged 14-16 (N=122), testing two interventions: (i) comparing AI-generated solutions with step-by-step explanations versus solutions alone and (ii) Socratic AI chatbots promoting critical thinking through incremental guidance versus non-Socratic AI offering direct solutions. Students completed school-related tasks under these conditions. We collected students' performance and interaction logs data, as well as their attitudes through self-reported surveys. Results indicated that AI-generated explanations significantly improved students' performance over solutions alone, highlighting the benefits of step-by-step guidance. The Socratic AI approach fostered significantly greater engagement and interaction. However, it did not achieve significant improvements in learning, and a higher fraction of students perceived it as less helpful. Furthermore, despite overall positive perceptions of AI assistance, students exhibited limited retention, failing to apply learned concepts to new situations without the aid of AI. These findings underscore both the potential and limitations of LLM-based chatbots in K-12 education, especially in balancing the development of critical thinking with user satisfaction.

### 2025-05-29 01-intro.Rmd
revised version

Integrating artificial intelligence (AI) into educational settings has been debated for over three decades [@roll2016evolution]. Recent advances in generative AI and large language model (LLM) chatbots have opened even more possibilities for improvements in educational practices [@yan2024promises;@kasneci2023chatgpt;@tlili2023if;@debets2025chatbots]. Although LLMs mimic human intelligence, their applications extend beyond imitation and include a range of academic tasks: facilitating problem-solving [@urban2024chatgpt], providing personalised assistance [@darvishi2024impact], language learning [@derakhshan2024chatgpt;@song2023enhancing], evaluation of students' work [@henkel2024can]. However, the impact of these applications on students' learning remains controversial, with proponents highlighting potential benefits and critics cautioning about risks and drawbacks [@farrokhnia2024swot;@weidinger2022taxonomy;@walczak2023challenges].

@@

The distinction between AI-generated *explanations* and *solutions* further allows us to evaluate whether providing explanations influences students' trust and judgment of AI-generated content. Indeed, LLMs can generate inaccurate solutions and flawed reasoning [@saxton2019analysing;@kalyan2021how;@hendrycks2021measuring]. However, there is evidence that providing explanations increases users' trust and acceptance of AI systems, especially when the reasoning behind the output is clear [@rai2020explainable;@ferrario2022explainability]. In an education setting, this may affect learning outcomes in a complex manner. Students may interpret well-written explanations as evidence that an answer is correct, even when it is not. Conversely, they may notice minor errors or inconsistencies in an explanation that reduces their trust, even if the overall solution is accurate. This situation motivates our second research question:

@@

Although classical pedagogical methods like the Socratic Method have long been valued for promoting critical thinking and more profound understanding [@lam2011socratic;@dalim2022promoting;@wilberding2021socratic;@cleveland2015beyond], there is a notable gap in the AI education literature regarding how to integrate these methods into AI-driven solutions. Existing AI tools, like ChatGPT, often focus on providing direct solutions or feedback, potentially limiting opportunities for active student engagement and reflection. Our study addresses this gap by conducting a parallel experimental intervention comparing Socratic-style AI interactions to more conventional, solution-focused AI interactions in a K-12 setting. This comparison contributes novel evidence on how different AI interaction modes can influence student learning processes and outcomes, informing the design of more pedagogically sound AI educational systems.

@@

A growing body of research highlights the tremendous potential of LLMs to improve student learning outcomes. A recent scoping review identifies 53 distinct use cases for LLMs in education [@yan2024practical], ranging from automatic grading and personalised feedback to teaching support and content generation, reflecting the high versatility and scalability of LLMs. Within this expanding literature, this study centers on *AI chatbots*, conversational agents powered by LLMs, and their educational applications, reflecting their increasing prominence and potential impact in this field.

Early investigations into the use of AI chatbots, such as ChatGPT, in educational settings have identified both opportunities and challenges [@farrokhnia2024swot;@noroozi2024generative]. On one hand, ChatGPT can serve as a personalised tutor, assist in drafting assignments, or generate creative prompts for classroom activities. On the other hand, concerns have been raised around academic integrity, overreliance on AI-generated content, and the need for critical thinking skills when interpreting chatbot outputs.

A recent review of over 70 studies offers a broad look at how chatbots are used in different areas of education [@debets2025chatbots]. It groups chatbot uses by roles like tutoring, giving feedback, and helping with administrative tasks, and examines how they are designed and used. The review finds that AI chatbots generally meet their goals. For example, a quasi-experimental study with 320 middle schoolers found that an AI teaching chatbot significantly improved math knowledge [@xing2025development]. Similarly, a meta-analysis of 24 RCTs shows that AI chatbots significantly enhance student outcomes, with stronger effects in higher education [@wu2024ai]. However, most studies are short-term, may suffer from publication bias, and often lack a strong theoretical foundation.

Despite promising results, key limitations remain. For instance, @er2024assessing found that students in a programming class performed better with instructor feedback, which was seen as more useful and fair than AI-generated feedback, highlighting the current gap between AI and human instruction. Similarly, evidence of performance boosts from non-education settings is mixed. A meta-analysis by @vaccaro2024combinations found that human-AI teams often performed worse than either humans or AI alone, suggesting a more nuanced relationship. In education, additional concerns are that integrating AI might encourage cheating, over-reliance, or superficial understanding [@farrokhnia2024swot;@yusuf2024generative;@eke2023chatgpt;@lee2024cheating], the risk of which is hard to evaluate without longitudinal studies.

@@

This study advances this literature on human-AI collaboration in educational contexts by examining the role of step-by-step reasoning and the impact of different modes of AI interaction. Additionally, it contributes to ongoing research exploring the factors that influence students' engagement with AI. Gaining insights into the value these combinations bring is crucial for developing practical guidelines for schools and educational institutions.

Most existing research focuses on general AI chatbots, like ChatGPT, which are not specifically designed for students and often lack clear pedagogical foundations. Our study highlights the importance of purposeful design and pedagogical goals, contributing to the growing field of AI chatbot design for education, or design for learning [@gavsevic2023empowering]. This research calls for a multidisciplinary approach combining pedagogy and human-computer interaction. Building on recent efforts [@dai2023can; @yan2024practical; @joksimovic2023opportunities], we examine how K–12 students interact with different AI chatbot styles. By doing so, the study underscores both the potential and challenges of applying classical pedagical methods, such as the Socratic approach, emphasizing the need to provide explanations rather than just answers.

>>>>

Integrating artificial intelligence (AI) into educational settings has been debated for over three decades [@roll2016evolution]. Recent advances in generative AI and large language model (LLM) chatbots have opened even more possibilities for improvements in educational practices [@yan2024promises;@kasneci2023chatgpt;@tlili2023if;@debets2025chatbots]. Although LLMs mimic human intelligence, their applications extend beyond imitation and include a range of academic tasks: facilitating problem-solving [@urban2024chatgpt], providing personalised assistance [@darvishi2024impact], language learning [@derakhshan2024chatgpt;@song2023enhancing], and evaluating students' work [@henkel2024can]. However, the impact of these applications on students' learning remains controversial, with proponents highlighting potential benefits and critics cautioning about risks and drawbacks [@farrokhnia2024swot;@weidinger2022taxonomy;@walczak2023challenges].

@@

The distinction between AI-generated *explanations* and *solutions* further allows us to evaluate whether providing explanations influences students' trust and judgment of AI-generated content. Indeed, LLMs can generate inaccurate solutions and flawed reasoning [@saxton2019analysing;@kalyan2021how;@hendrycks2021measuring]. However, there is evidence that providing explanations increases users' trust and acceptance of AI systems, especially when the reasoning behind the output is clear [@rai2020explainable;@ferrario2022explainability]. In an educational setting, this may affect learning outcomes in a complex manner. Students may interpret well-written explanations as evidence that an answer is correct, even when it is not. Conversely, they may notice minor errors or inconsistencies in an explanation that reduce their trust, even if the overall solution is accurate. This situation motivates our second research question:

@@

Although classical pedagogical methods, such as the Socratic Method, have long been valued for promoting critical thinking and a deeper understanding [@lam2011socratic;@dalim2022promoting;@wilberding2021socratic;@cleveland2015beyond], there is a notable gap in the AI education literature regarding how to integrate these methods into AI-driven solutions. Existing AI tools, such as ChatGPT, often focus on providing direct solutions or feedback, which can potentially limit opportunities for active student engagement and reflection. Our study addresses this gap by conducting a parallel experimental intervention that compares Socratic-style AI interactions with more conventional, solution-focused AI interactions in a K-12 setting. This comparison contributes novel evidence on how different AI interaction modes can influence student learning processes and outcomes, informing the design of more pedagogically sound AI educational systems.

@@

A growing body of research highlights the tremendous potential of LLMs to improve student learning outcomes. A recent scoping review identifies 53 distinct use cases for LLMs in education [@yan2024practical], ranging from automatic grading and personalised feedback to teaching support and content generation, reflecting the high versatility and scalability of LLMs. Within this expanding literature, this study centres on *AI chatbots*, conversational agents powered by LLMs, and their educational applications, reflecting their increasing prominence and potential impact in this field.

Early investigations into the use of AI chatbots, such as ChatGPT, in educational settings have identified both opportunities and challenges [@farrokhnia2024swot;@noroozi2024generative]. On the one hand, ChatGPT can serve as a personalised tutor, assist in drafting assignments, or generate creative prompts for classroom activities. On the other hand, concerns have been raised around academic integrity, overreliance on AI-generated content, and the need for critical thinking skills when interpreting chatbot outputs.

A recent review of over 70 studies provides a comprehensive overview of how chatbots are used in various educational contexts [@debets2025chatbots]. It classifies chatbot applications by roles, including tutoring, giving feedback, and helping with administrative tasks, and examines how they are designed and utilised. The review finds that AI chatbots generally meet their goals. For example, a quasi-experimental study with 320 middle schoolers found that an AI teaching chatbot significantly improved math knowledge [@xing2025development]. Similarly, a meta-analysis of 24 RCTs shows that AI chatbots significantly enhance student outcomes, with stronger effects in higher education [@wu2024ai]. However, most studies are short-term, may be subject to publication bias, and often lack a strong theoretical foundation.

Despite promising results, key limitations remain. For instance, @er2024assessing found that students in a programming class performed better with instructor feedback, which was seen as more useful and fair than AI-generated feedback, highlighting the current gap between AI and human instruction. Similarly, evidence of performance boosts from non-education settings is mixed. A meta-analysis by @vaccaro2024combinations found that human-AI teams often performed worse than either humans or AI alone, suggesting a more nuanced relationship. In education, additional concerns are that integrating AI might encourage cheating, over-reliance, or superficial understanding [@farrokhnia2024swot;@yusuf2024generative;@eke2023chatgpt;@lee2024cheating], a risk that is difficult to evaluate without longitudinal studies.

@@

This study advances the literature on human-AI collaboration in educational contexts by examining the role of step-by-step reasoning and the impact of different modes of AI interaction. Additionally, it contributes to ongoing research exploring the factors that influence students' engagement with AI. Gaining insights into the value these combinations bring is crucial for developing practical guidelines for schools and educational institutions.

Most existing research focuses on general AI chatbots, like ChatGPT, which are not specifically designed for students and often lack clear pedagogical foundations. Our study highlights the importance of purposeful design and pedagogical goals, contributing to the growing field of AI chatbot design for education or design for learning [@gavsevic2023empowering]. This research calls for a multidisciplinary approach combining pedagogy and human-computer interaction. Building on recent efforts [@dai2023can; @yan2024practical; @joksimovic2023opportunities], we examine how k–12 students interact with different AI chatbot styles. By doing so, the study underscores both the potential and challenges of applying classical pedagogical methods, such as the Socratic approach, emphasizing the need to provide explanations rather than just answers.

### 2025-05-29 02-methods.Rmd
revised version

- **User authentication**: A secure and anonymous login system for students.
- **Web Survey**: A web interface to collect data from students, including answering survey questions and perform simple tasks, such as writing a short essay or solving simple math problems.
- **AI Chatbot**: A chatbot powered by GPT-4 provided students with personalised assistance and tutoring during the session. A new chatbot instance was associated with each question for each student, isolating each conversation from interactions in previous questions. The chatbot was available to students only for specific questions or tasks under the researchers' control.
- **Data logging**: A SQL database storing students' conversation logs with the AI chatbot, including texts, timestamps, and survey responses. 
- **Randomised assignment**: A system to randomly assign registered students to different versions of the AI chatbot or treatment groups to explore the causal impact of various configurations on students' interactions and performance.

@@

We recruited students after receiving ethical approval from the European Commission's internal Ethical Review Board, ensuring that the consent procedures, data protection requirements, and the experimental protocol complied with local laws and ethical research standards. The data privacy protection protocol was approved by the Commission's data protection officer. As an additional safeguard, given the minor age of the participants, we required their parents or legal guardians to provide us with written consent, allowing their children to participate in the study. The students also had to give their assent to participate by completing an online consent form.

@@

[^jar-gpt4]: Using GPT 4.0, we uploaded the image of the coin jar asking for an estimate of the value of coins for ten times. We then selected the median response for the experiment.

@@

[^prompts]: An AI's system message guides how the AI interprets the conversations by setting paramters for interaction.

```{r prompts}
df_prompts <- data.frame(
  "AI Tutor" = c("Socratic", "Non-Socratic"),
  "Prompt instruction" = c("You are a Socratic tutor. You always answer using the Socratic style, asking just the right questions to help students learn to think for themselves, breaking down the problem into simpler parts until it's at the right level for them. You provide concise information and explanations understandable for 8th to 10th grade students.", "You are a didactic tutor. You provide concise information and explanations understandable for 8th to 10th grade students.") # nolint
)
kbl(
  df_prompts, booktabs = TRUE,
  caption = "Prompt instructions associated with the AI Tutor's treatments"
) |>
  kableExtra::column_spec(2, "4in")
```

This setup allows us to investigate the impact of different pedagogical AI tutoring approaches on students' performance and perceptions. Specifically, we asked students to use the AI tutor while performing three tasks: guess an unknown quantity ("How much water in litres do students consume at our school each week?"); express an opinion and write a short essay ("What is your opinion about the effect of social media on teenagers?"); respond to physics questions on how sound propagates in different media ("In which of the following materials does sound travel faster?"). These questions enabled us to assess the AI tutor's impact on students' learning and problem-solving abilities.

We focused on two primary metrics: (1) confidence in their answers, (2) perceived usefulness of interacting with the AI tutor. Specifically,  students were asked, "How confident are you that the answer you provided is accurate?" on a five-point scale ranging from "not confident at all" to "very confident." They were also asked "How helpful was it to interact with the AI tutor?" This was also rated on a five-point scale, from "Not at all helpful" to "Very helpful." Only for task 3, we had an additional metric which was the correctness of their answers.

@@

To examine student interactions with AI and their potential impact on learning and critical thinking, we collected a range of self-reported and behavioral data. The full questionnaire is provided in the Appendix (Section \@ref(sec:questionnaire)).

@@

To better understand students' general attitudes toward AI, we selected four questions from Schepman and Rodway's AI Attitudes Scale [@schepman2020initial]. These questions explored students; beliefs about AI's societal benefits, potential dangers, capacity to support learning, and likelihood of misuse by students (Appendix, Section \@ref(sec:questionnaire)). We also asked students about their direct experience using ChatGPT for homework --- the most common chatbot at the time of the study --- and their beliefs of how many of their peers use ChatGPT for their homework, thereby capturing both individual behaviour and the perceived social norm around AI usage. These measures help establish baseline orientations that may moderate how students engage with and evaluate AI explanations (related to RQ-1) and how they perceive different AI interaction styles (related to RQ-3).

@@

Academic skills or habits may moderate the effect of AI interactions on problem solving activities. Accordingly, to acconut for participants' heterogeneity in academic skills, we asked students to report their average school grades within five categories. We also asked students about their academic habits or the challenges they face at school, such as how often they complete their homework assignments on time and what factors affect their ability to do so.

@@

To quantify engagement, we analyzed students' interaction logs with the AI Tutor, recording the number of turns and the word counts as a proxy for interaction intensity. While this provides a basic behavioral metric, it does not capture the quality of cognitive engagement or helpfulness, limiting its direct relevance to RQ-1 and RQ-2. Therefore, we also asked students about how useful they found the AI interactions and how confident they were in their answers to selected tasks. These self-reported metrics are needed to assess whether explanations were helpful to students (for RQ-1) and whether Socratic dialogue promoted critical thinking or confidence (for RQ-2).

>>>>

- **User authentication**: A secure and anonymous login system for students.
- **Web Survey**: A web interface to collect data from students, including answering survey questions and performing simple tasks, such as writing a short essay or solving simple math problems.
- **AI Chatbot**: A chatbot powered by GPT-4 provided students with personalised assistance and tutoring during the session. A new chatbot instance was associated with each question for each student, isolating each conversation from interactions in previous questions. The chatbot was available to students only for specific questions or tasks under the researchers' control.
- **Data logging**: A SQL database storing students' conversation logs with the AI chatbot, including texts, timestamps, and survey responses. 
- **Randomised assignment**: A system to randomly assign registered students to different versions of the AI chatbot or treatment groups to explore the causal impact of various configurations on students' interactions and performance.

@@

We recruited students after receiving ethical approval from the European Commission's internal Ethical Review Board, ensuring that the consent procedures, data protection requirements, and experimental protocol complied with local laws and ethical research standards. The data privacy protection protocol was approved by the Commission's data protection officer. As an additional safeguard, given the minor age of the participants, we required their parents or legal guardians to provide us with written consent, allowing their children to participate in the study. The students also had to give their assent to participate by completing an online consent form.

@@

[^jar-gpt4]: Using GPT 4.0, we uploaded the image of the coin jar asking for an estimate of the value of coins ten times. We then selected the median response for the experiment.

@@

[^prompts]: An AI's system message guides how the AI interprets the conversations by setting parameters for interaction.

```{r prompts}
df_prompts <- data.frame(
  "AI Tutor" = c("Socratic", "Non-Socratic"),
  "Prompt instruction" = c("You are a Socratic tutor. You always answer using the Socratic style, asking just the right questions to help students learn to think for themselves and breaking down the problem into simpler parts until it's at the right level for them. You provide concise information and explanations understandable for 8th to 10th grade students.", "You are a didactic tutor. You provide concise information and explanations understandable for 8th to 10th grade students.") # nolint
)
kbl(
  df_prompts, booktabs = TRUE,
  caption = "Prompt instructions associated with the AI Tutor's treatments"
) |>
  kableExtra::column_spec(2, "4in")
```

This setup allows us to investigate the impact of different pedagogical AI tutoring approaches on students' performance and perceptions. Specifically, we asked students to use the AI tutor while performing three tasks: guess an unknown quantity ("How much water in litres do students consume at our school each week?"), express an opinion and write a short essay ("What is your opinion about the effect of social media on teenagers?"); respond to physics questions on how sound propagates in different media ("In which of the following materials does sound travel faster?"). These questions enabled us to assess the AI tutor's impact on students' learning and problem-solving abilities.

We focused on two primary metrics: (1) confidence in their answers and (2) perceived usefulness of interacting with the AI tutor. Specifically,  students were asked, "How confident are you that the answer you provided is accurate?" on a five-point scale ranging from "not confident at all" to "very confident." They were also asked, "How helpful was it to interact with the AI tutor?" This was also rated on a five-point scale, from "Not at all helpful" to "Very helpful." Only for task 3, we had an additional metric, which was the correctness of their answers.

@@

To examine student interactions with AI and their potential impact on learning and critical thinking, we collected a range of self-reported and behavioural data. The full questionnaire is provided in the Appendix (Section \@ref(sec:questionnaire)).

@@

To better understand students' general attitudes toward AI, we selected four questions from Schepman and Rodway's AI Attitudes Scale [@schepman2020initial]. These questions explored students; beliefs about AI's societal benefits, potential dangers, capacity to support learning, and the likelihood of misuse by students (Appendix, Section \@ref(sec:questionnaire)). We also asked students about their direct experience using ChatGPT for homework --- the most common chatbot at the time of the study --- and their beliefs about how many of their peers use ChatGPT for their homework, thereby capturing both individual behaviour and the perceived social norm around AI usage. These measures help establish baseline orientations that may moderate how students engage with and evaluate AI explanations (related to RQ-1) and how they perceive different AI interaction styles (related to RQ-3).

@@

Academic skills or habits may moderate the effect of AI interactions on problem-solving activities. Accordingly, to account for participants' heterogeneity in academic skills, we asked students to report their average school grades within five categories. We also asked students about their academic habits or the challenges they face at school, such as how often they complete their homework assignments on time and what factors affect their ability to do so.

@@

To quantify engagement, we analyzed students' interaction logs with the AI Tutor, recording the number of turns and the word counts as a proxy for interaction intensity. While this provides a basic behavioural metric, it does not capture the quality of cognitive engagement or helpfulness, limiting its direct relevance to RQ-1 and RQ-2. Therefore, we also asked students about how useful they found the AI interactions and how confident they were in their answers to selected tasks. These self-reported metrics are needed to assess whether explanations were helpful to students (for RQ-1) and whether Socratic dialogue promoted critical thinking or confidence (for RQ-2).

### 2025-05-29 04-discussion.Rmd
edits
revised version

The present study aims to contribute to our understanding of the impact of pedagogically aligned configurations of an LLM-based tool on high-school students’ learning and attitudes with a special focus on their critical thinking.

With the rapid adoption of AI in education, the focus on students' competencies and skills, such as critical thinking, has become particularly urgent [@walter2024embracing;@holmes2023guidance]. Despite the rapidly increasing body of research on the impact of LLMs on students’ learning, most studies focus on participants in higher education [e.g.,@suriano2025student], which do not necessarily generate insights that apply to younger students who have different cognitive maturity. In addition, despite the increasing research interest regarding the impact of LLMs on students’ learning, there is little consensus among researchers, especially due to students’ over-reliance on AI tools [@zhai2024effects].

<!-- as individuals increasingly favor fast and optimal solutions over slow ones constrained by practicality -->

The results of this study focus on the role of two interventions on the configuration of an LLM chatbot for educational support: (i) step-by-step explanations and (ii) the Socratic method with guiding questions. First, we found that students significantly benefit from AI-generated step-by-step reasoning accompanying solutions, mainly when performing open-ended tasks like estimating unknown quantities. This finding aligns with expectations of current research on the use of LLM chatbots and their ability to provide explanations for learning purposes [@wu2024critical]. It also aligns with previous research demonstrating that LLMs can enhance student performance [@xing2025development]. However, our results underscore that the key mechanism driving AI improvements is the step-by-step reasoning provided by the AI, which allows students to understand better and engage with problem-solving.

<!-- This insight suggests that teachers should focus on educating students to enhance their ability to evaluate and judge the correctness of AI-generated reasoning. -->

Furthermore, our study reveals that step-by-step reasoning not only helps students in solving problems but also enhances students' ability to evaluate AI-generated information critically. Critical thinking is indeed a complex cognitive process that involves evaluation and analysis of a given information [@lai2011critical]. The mobilisation of students' critical thinking in the specific task was evidenced by the more positive evaluations of human-generated guesses compared to AI-generated solutions, suggesting that students could better assess and challenge AI predictions. This contribution extends the existing literature by showing that AI can foster critical thinking and analytical skills when coupled with transparent reasoning, instead of being presented  as a "black box" [@bearman2023learning].

In addition, we compared various ways to structure student-AI interactions and, more specifically, the effectiveness of Socratic AI (an interactive, questioning-based AI) with non-Socratic AI. Our results showed that Socratic AI was more engaging and promoted greater interaction, aligning with the notion that AI-student interactions should be dynamic and dialogical [@suriano2025student]. However, contrary to our expectations, we found no significant differences in students' self-reported confidence in the accuracy of their answers or in the correctness of their responses despite the more frequent interactions with the Socratic AI. Additionally, the Socratic AI's perceived helpfulness was rated lower compared to the non-Socratic AI. These results cast some doubts on the effectiveness of Socratic AI in short-term tasks or, more broadly, the effect of certain kinds of AI-student interactions. As such, existing pedagogical practices extensively used in human-human interaction might not always work in human-AI interaction. In addition, while recent research attempts to understand the quality of Socratic LLMs for critical thinking [e.g.,@favero2024enhancing], these studies use simulations without experiments with human participants. Our results underscore the need for adapting existing pedagogical paradigms or proposing new ones for integrating LLM tools into pedagogical practices effectively.

Our findings also indicate that simply interacting with AI cannot promote meaningful and lasting learning. In our initial test on sound propagation, approximately half of the students could revise their initial answers based on AI interactions, improving their response accuracy from 30% to 70%. However, in a verification task where students had no access to AI, the majority failed to identify the correct answer, mistakenly claiming that sound propagates faster in water than in gold. Most students exhibited this misunderstanding, which supports key concerns that AI-generated answers alone do not scaffold effective learning, and the mere implementation of Socratic AI does not mitigate this risk.

Our results suggest that AI has great potential as an educational tool, but its implementation requires careful considerations. First, students with prior experience in using AI tools may not readily adopt new pedagogical approaches, especially if they perceive them as less effective than commercially available alternatives. Second, the effectiveness of the pedagogical approach may vary depending on the nature of the task, complicating the design of a one-size-fits-all solution for AI-assisted learning.

Several limitations of our study should be acknowledged. The small sample size limits the generalizability of our findings, though the controlled environment and the use of multiple tasks help mitigate potential noise in the data. Also, our experiment focused on short-term results. Although short term results are important to foster adoption, it is unclear the effectiveness of AI in the long-term. Another limitation is that learning retention was tested using only one specific physics question. Although we controlled for prior knowledge and carefully designed a simple task to fit within a 40-minute intervention, additional questions would be needed to rule out confounding factors. Additionally, our study combined objective performance metrics with self-assessed ratings of helpfulness and confidence. However, individual perceptions do not necessarily reflect actual learning [@noroozi2025does]. Further research is needed to validate our findings using objective learning measures.

Finally, we conducted our study with a cohort of students possessing strong English proficiency and a clear understanding of the limitations of AI, which may not be representative of the general student population. Additionally, we focused on only two schools, which allowed us to control for school-specific fixed effects. However, we recognize that the impact of the treatments may differ across schools, and a broader investigation involving many more institutions would be necessary to explore this variability. Furthermore, the experiment was carried out at school and in a secure and anonymous digital environment, with a robust protocol developed to ensure the safe and ethical handling of AI-based interactions in experimental settings. However, it remains to be seen if the results of our analysis will remain when students use AI in the field.

@@

While our study focused on comparing a fairly general pedagogical approach---Socratic AI, future research should explore this pedagogical method in a more nuanced way and consider alternative pedagogical approaches to facilitate scalability of AI tools across diverse tasks and educational contexts. Yet, our findings have important implications for designing AI tutors, highlighting the importance of  providing transparent AI-generated step-by-step reasoning and the challenges of fostering learning through guided AI-student interactions. Therefore, AI systems must engage students interactively and provide learning opportunities for students' critical thinking and problem-solving skills to maximise their educational value. Future developments should focus on refining the integration of AI-generated reasoning and ensuring that AI tools are adaptable to various learning tasks and students' needs.

>>>>

This study contributes to our understanding of the impact of pedagogically-aligned configurations of an LLM-based tool on the learning and attitudes of high school students, with a special focus on their critical thinking.

## Overview of the findings

With the rapid adoption of AI in education, the focus on students' competencies and skills, such as critical thinking, has become particularly urgent [@walter2024embracing;@holmes2023guidance]. Despite the rapidly increasing body of research on the impact of LLMs on students’ learning, most studies focus on participants in higher education [e.g.,@suriano2025student], which may not generate insights applicable to younger students with different cognitive maturity. In addition, despite the increasing research interest regarding the impact of LLMs on students’ learning, there is little consensus among researchers, mainly due to students’ over-reliance on AI tools [@zhai2024effects].

The results of this study focus on the role of two interventions in configuring an LLM chatbot for educational support: (i) step-by-step explanations and (ii) the Socratic method with guiding questions. First, we found that students significantly benefit from AI-generated step-by-step reasoning accompanying solutions, mainly when performing open-ended tasks like estimating unknown quantities (RQ1). This finding aligns with expectations of current research on the use of LLM chatbots and their ability to provide explanations for learning purposes [@wu2024critical]. It also aligns with previous research demonstrating that LLMs can enhance student performance [@xing2025development]. However, our results underscore that the key mechanism driving AI improvements is the step-by-step reasoning provided by the AI, which allows students to understand better and engage with problem-solving.

Furthermore, our study reveals that step-by-step reasoning not only helps students in solving problems but also enhances students' ability to evaluate AI-generated information critically (RQ2). Critical thinking is indeed a complex cognitive process that involves the evaluation and analysis of a given information [@lai2011critical]. The mobilisation of students' critical thinking in the specific task was evidenced by the more positive evaluations of human-generated guesses compared to AI-generated solutions, suggesting that students could better assess and challenge AI predictions. This contribution extends the existing literature by showing that AI can foster critical thinking and analytical skills when coupled with transparent reasoning rather than being presented  as a "black box" [@bearman2023learning].

In addition, we compared various ways to structure student-AI interactions and, more specifically, the effectiveness of Socratic AI (an interactive, questioning-based AI) with non-Socratic AI. Our results showed that Socratic AI was more engaging and promoted greater interaction, aligning with the notion that AI-student interactions should be dynamic and dialogical [@suriano2025student]. However, contrary to our expectations, we found no significant differences in students' self-reported confidence in the accuracy of their answers or in the correctness of their responses despite the more frequent interactions with the Socratic AI (RQ4). Additionally, the Socratic AI's perceived helpfulness was rated lower compared to the non-Socratic AI (RQ5). These results cast some doubts on the effectiveness of Socratic AI in short-term tasks or, more broadly, the effect of certain kinds of AI-student interactions. As such, existing pedagogical practices extensively used in human-human interaction might not always work in human-AI interaction. In addition, while recent research attempts to understand the quality of Socratic LLMs for critical thinking [e.g.,@favero2024enhancing], these studies use simulations without experiments with human participants. Our results underscore the need for adapting existing pedagogical paradigms or proposing new ones for integrating LLM tools into pedagogical practices effectively.

Interestingly, our findings indicate that simply interacting with AI cannot promote meaningful and lasting learning (RQ3). In our initial test on sound propagation, approximately half of the students could revise their initial answers based on AI interactions, improving their response accuracy from 30% to 70%. However, in a verification task where students had no access to AI, the majority failed to identify the correct answer, mistakenly claiming that sound propagates faster in water than in gold. Most students exhibited this misunderstanding, which supports key concerns that AI-generated answers alone do not scaffold effective learning and the mere implementation of Socratic AI does not mitigate this risk.

Our results suggest that AI has great potential as an educational tool, but its implementation requires careful consideration. First, students with prior experience in using AI tools may not readily adopt new pedagogical approaches, especially if they perceive them as less effective than commercially available alternatives. Second, the effectiveness of the pedagogical approach may vary depending on the nature of the task, complicating the design of a one-size-fits-all solution for AI-assisted learning.

## Limitations of the study and future work

Several limitations of our study should be acknowledged. The small sample size limits the generalizability of our findings, though the controlled environment and the use of multiple tasks help mitigate potential noise in the data. Also, our experiment focused on short-term results. Although short-term results are important to foster adoption, it is unclear the effectiveness of AI in the long term. Another limitation is that learning retention was tested using only one specific physics question. Although we controlled for prior knowledge and carefully designed a simple task to fit within a 40-minute intervention, additional questions would be needed to rule out confounding factors. Additionally, our study combined objective performance metrics with self-assessed ratings of helpfulness and confidence. However, individual perceptions do not necessarily reflect actual learning [@noroozi2025does]. Further research is needed to validate our findings using objective learning measures.

Finally, we conducted our study with a cohort of students possessing strong English proficiency and a clear understanding of the limitations of AI, which may not be representative of the general student population. Additionally, we focused on only two schools, which allowed us to control for school-specific fixed effects. However, we recognise that the impact of the treatments may differ across schools, and a broader investigation involving many more institutions would be necessary to explore this variability. Furthermore, the experiment was carried out at school and in a secure and anonymous digital environment, with a robust protocol developed to ensure the safe and ethical handling of AI-based interactions in experimental settings. However, it remains to be seen if the results of our analysis will remain when students use AI in the field.

@@

While our study focused on comparing a fairly general pedagogical approach---Socratic AI, future research should explore this pedagogical method in a more nuanced way and consider alternative pedagogical approaches to facilitate the scalability of AI tools across diverse tasks and educational contexts. Yet, our findings have important implications for designing AI tutors, highlighting the importance of  providing transparent AI-generated step-by-step reasoning and the challenges of fostering learning through guided AI-student interactions. Therefore, AI systems must engage students interactively and provide learning opportunities for students critical thinking and problem-solving skills to maximise their educational value. Future developments should focus on refining the integration of AI-generated reasoning and ensuring that AI tools are adaptable to various learning tasks and students' needs.

### 2025-05-29 main_report.Rmd
edits
revised version

---
title: |
  Exploring the Role of AI Chatbots in K-12 Education: A Comparative Study of Socratic and Non-Socratic Approaches

author:
  - Andrea Blasco:
      institute:
        - jrc
      correspondence: "yes"
      email: andrea.blasco@ec.europa.eu
  - Vicky Charisi:
      institute: [jrc, harvard]
      correspondence: "yes"
      email: vcharisi@law.harvard.edu

institute:
  - jrc: European Commission, Joint Research Centre, Brussels, Belgium
  - harvard: Harvard University, Berkman Klein Center, Cambridge, MA, USA

@@

header-includes: |
  ```{=latex}
  % Break sections into new page
  \usepackage{titlesec}
  \newcommand{\sectionbreak}{\clearpage}

>>>>

---
params:
  anonymous: false

title: |
  AI Chatbots in K-12 Education: An Experimental Study of Socratic vs. Non-Socratic Approaches and the Role of Step-by-Step Reasoning

author:
  - "`r ifelse(!params$anonymous, 'Andrea Blasco', 'Anonymous') `":
      institute:
        - jrc
      correspondence: "yes"
      email: "`r ifelse(!params$anonymous, 'andrea.blasco@ec.europa.eu', 'Email') `"
  - "`r ifelse(!params$anonymous, 'Vicky Charisi', 'Anonymous') `":
      institute: [jrc, harvard]
      correspondence: "yes"
      email: "`r ifelse(!params$anonymous, 'vcharisi@law.harvard.edu', 'Email') `"

institute:
  - jrc: "`r ifelse(!params$anonymous, 'European Commission, Joint Research Centre, Brussels, Belgium', 'Anonymous Institute') `"
  - harvard: "`r ifelse(!params$anonymous, 'Harvard University, Berkman Klein Center, Cambridge, MA, USA', 'Anonymous Institute')`"

@@

header-includes: |
  ```{=latex}
  % Break sections into new page
  \usepackage{titlesec}
  \newcommand{\sectionbreak}{\clearpage}
  \usepackage{soul} % \hl{} = highlights
